Modelling BCG vaccination in the UK

### 2019-12-11 rmds/09-model-fitting.Rmd
complete dodd typos

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding its structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occurring in England. In order to use this model to understand TB transmission, and the impact of different BCG vaccination policies (Chapter \@ref(model-results)) this uncertainty needs to be reduced and the parameter space tightened to reflect more realistic ranges. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantitative way "better" than other parameter sets. An alternative to model fitting is using a model parameterised with expert knowledge only for inference. This approach is not appropriate here due to the large amount of uncertainty for many of the model parameters. Any inference based on just the parametisation from the previous chapter would have large credible intervals, reflect reality poorly, and likely be biased in multiple areas.

This chapter details an approach to fitting a infectious disease model to data using the state-space model formulation and Bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad-hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000. A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, this may result in over-fitting and so has not been implemented here.

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging from tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates.[@Murray2015] Unfortunately, many of these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to fit the previously detailed state space model.

@@

 * *Re-sampling:* The particle stocke is restorted to equal weights by re-sampling particles, with replacement, with the probability of each sample being drawn being proportional to its weight.

@@

    *  *Re-sampling:* The parameter particle stock is restored to equal weights by re-sampling particles, with replacement, with the probability of each sample being drawn being proportional to its weight.

@@

The accuracy of the marginal likelihood estimate returned by the particle filter is dependent on the number of particles used, the number of observed data points, the parameter sample, and the complexity of the model. As the number of particles tends towards infinity the likelihood estimate provided by the particle filter should tend towards the exact solution. This suggests that choosing a very high number of particles may be the optimal solution in terms of accuracy. Unfortunately, each particle requires a full model simulation, which for complex models can be computationally costly. This means that using very large numbers of particles is not tractable. For this reason it is necessary to determine an optimal number of particles that both provides an adequately accurate estimate of the likelihood whilst being computationally tractable.

@@

The DIC has two limitations. The first of these is that in its derivation it is assumed that the model that generates future observations encompasses the true model. This assumption may not hold in all circumstances. The second limitations is that the observed data is used to construct both the posterior distribution and to estimate the DIC. This means that the DIC tends to select for over-fitted model.[@Gelman:nll_LBlw]

@@

Where $\operatorname {cov}$  is the co-variance, $\sigma _{X}$ is the standard deviation of $X$, and $\sigma_Y$  is the standard deviation of $Y$. Where $X$ is the input and $Y$ is the output.

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC-SMC algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvenation steps in all scenarios evaluated. This resulted in spuriously tight posterior distributions. Ad-hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and indicative only.

### Ad-hoc calibration

Minor alterations to the model fitting pipeline had little impact on the quality of the fit. To attempt to improve the quality of the model fit I used a combination of ad-hoc approaches. As a first step I introduced a calibration model with variation allowed only in the fitted parameters with all other parameters using point estimates. This allowed a reduced number of particles to be used to estimate the marginal likelihood and hence dramatically decreased compute cost and run-times. This reduced model was then used for the following tests:

@@

[@Murray2015] suggests that increasing the number of particles used in SMC may improve the quality of the model fit. Unfortunately as `LibBi` stores SMC particle paths in Random access memory (RAM) the number of particles was restricted. An additional limitation is that the current rejuvenation step also need to be stored in RAM. Attempts to increase the amount of available RAM (64 GB) using a 500GB SWAP (virtual memory) drive increased the upper limit on the number of particles but gains from this were restricted due to thrashing^[Thrashing: https://en.wikipedia.org/wiki/Thrashing_(computer_science)]. Thrashing occurs when to much data is written to SWAP memory in a short period of time and usually results in a system crash.  A major cause of the high RAM requirements of the model was that `LibBi` stores all parameters defined as initial conditions across every time point in the model, regardless of the settings used. Attempting to reduce the RAM footprint by fitting a greater number of parameters allowed for a greater number of particles to be used but also increased the model degrees of freedom and hence upped the required number of particles required by a greater amount than provided by fitting the parameter. Reducing the number of particles used to estimate the marginal likelihood allowed a greater number of particles to be used in the outer SMC step but this resulted in highly inaccurate estimates of the likelihood. These highly inaccurate likelihood estimates resulted in the SMC-SMC algorithm focusing on parameter sets that had a low likelihood estimate yet fit the data poorly - ultimately leading to poorer model fits.

Varying the proposal size, rejuvenation threshold, and number of rejuvenation steps showed some promise at improving the quality of the model fit but ultimately computational constraints limited how much progress could be made using this approach. It is possible that a much longer run time could result in an improved fit to data with no other model changes. Alternatively these results may be driven by the model being to complex for an SMC-SMC approach to be viable.

@@

Whilst none of the scenarios fitted the data well, scenarios that included variable transmission probability in young adults fitted the data much better than those that did not (Table \@ref(tab:09-scenarios-table)). When considered on its own, allowing non-UK born mixing to vary for young adults had only a small impact on the quality of the model fit in comparison to allowing the transmission probability to vary. However, the scenario that allowed both the transmission probability and non-UK born mixing to vary fit the data much better than any other scenario considered (Table \@ref(tab:09-scenarios-table)). In the following sections only the results from this scenario will be discussed.

@@

```{r 09-table-incidence-preds, message = FALSE, warning = FALSE}
sum_obs_table %>% 
  set_names(., nm = c(names(.)[-ncol(.)], "Predicted Incidence (95% CI)")) %>% 
   set_names(., nm = names(.) %>% str_replace_all("Incidence", "Cases")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Observed versus predicted overall TB cases for years that the model was fitted to.", 
  caption = "Observed versus predicted overall TB cases for years that the model was fitted to. (95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  {.}
```

@@

```{r 09-prior-posterior, warning = FALSE, message = FALSE}
sum_prior_posterior %>% 
  set_names(., nm = . %>% 
              str_replace("or", "or (95\\\\%CI)")) %>%
  mutate(Parameter = latex_param) %>% 
  select(-latex_param) %>% 
kable(
  caption.short = "Prior and posterior medians for fitted model parameters", 
  caption = "Prior and posterior medians for fitted model parameters. (95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.",
  booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(latex_options = c("HOLD_position")) %>% 
  {.}
```

@@

In this chapter I have formulated the disease transmission model developed in the previous chapter as a state-space model, developed a model fitting pipeline to fit this model to observed data, discussed the approaches taken to try and improve the quality of the model fit, and presented preliminary results from the model fitting pipeline. The model fit the observed data poorly using the approach layed out here. Whilst multiple ad-hoc approaches were used to try and improve the quality of the model fit these did little to improve it. The model consistently under-predicted overall TB incidence and also failed to reproduce the observed age distribution. There was little evidence of a good fit to trends over time in the observed data. Model comparison showed that models that included modifying parameters for the transmission probability, and non-UK born mixing in young adults fitted the observed data better than those that did not. The estimated posterior distributions for all parameters were spuriously tight in comparison to the prior distributions. This may be evidence of poor mixing meaning that any conclusions drawn about the parameter sets selected may be incorrect. Parameters that contributed to recent TB transmission dominated those that contributed across the time-span of the model in the parameter sensitivity analysis.

As discussed earlier in this chapter, SMC-SMC was used for all model fitting. This uses an external SMC step to estimate the posterior distribution as well as an internal SMC step to estimate the marginal likelihood. In general, Bayesian model fitting approaches are beneficial as they allow prior information to be fully incorporated into model fitting and they produce posterior distribution estimates rather than point estimates.[@Murray2015; @Toni2009] There are two main families of approaches, SMC and MCMC. Using SMC to estimate the posterior distribution has numerous advantages over MCMC. The first of these is that MCMC approaches are sensitive to their initial conditions. If a model has multiple local best fits MCMC may only converge to a single minima rather than fully exploring the posterior distribution. Multiple MCMC chains may be used to try and account for this but as each chain must be independently run to convergence only a few concurrent chains are likely to be practical. SMC on the other hand is initialised with a large sample from the prior distribution, meaning that local minimas are more likely to be explored. Parameter particle weighting and re-sampling then balances the contribution to the posterior distribution of these local minimas based on their fit to the observed data. Additionally, MCMC approaches are by definition sequential,[@Murray2015] although if they make use of particle filters these can be run in parallel. Increasing the number of particles in a filter may lead to an increase in the chain mixing rate of the MCMC chain but as particles numbers are increased any returns will decrease. To account for this multiple chains are often used, but as outlined above the burn-in required for each chain limits the potential speed-up. In comparison, each SMC parameter particle can have it's marginal likelihood computed separately. Although the re-sampling step remains a bottleneck as it can only be completed once all marginal likelihoods have been computed. On the other hand SMC is less interactive than MCMC meaning that model fitting is harder to inspect when it is in progress. This is because SMC is not sequential, unlike an MCMC run for which each draw can be inspected as it is computed. Similarly, as SMC is not a sequential technique multiple runs cannot be combined. This means that model fitting must be done in a single run using a priori knowledge to judge the number of MCMC rejuvenation steps required, and the expected total run time. SMC will also have a variable run time based on the effective sample size as rejuvenation only happens when parameter particles have been depleted beyond a certain point. An additional benefit of SMC approaches are that they can theoretically be extended to model selection as well as parameter posterior estimation - effectively estimating posteriors for candidate model structures.[@Toni2009] Beyond SMC and MCMC there are multiple other model fitting algorithms that each have their own strengths and benefits - discussion of these is outside the scope of this thesis.

The particle filter has been shown to provide an unbiased estimate of the likelihood for arbitrary state-space models.[@Murray2015] As a full information technique the particle filter provides a more accurate estimate of the likelihood than other approximate techniques with relatively little tuning or user interaction.[@Murray2015] The major downside of the particle filter, compared to other approaches, is the high compute requirements, with each particle requiring a full model simulation. For highly complex models, the particle filter approach may not be tractable or a reduced level of accuracy of the marginal likelihood estimate must be accepted. In addition, the bootstrap particle filter may become depleted (i.e variance between particles is reduce to such an extent that the effective sample size becomes small) when the model being fitted is a poor fit for the observed data.[@Murray2015] Whilst this can be resolved using additional particles, or rejuvenation steps, this may not be computationally tractable. There are several alternative particle filters that seek to mitigate these issues but many of them are highly complex and do not significantly reduce the required compute - see [@Murray2015] for a detailed discussion of some of these alternatives. An alternative approach to the particle filter is to use an approximate technique, such as approximate Bayesian computation (ABC). ABC can be used to estimate the likelihood by comparing observed and simulated data.[@Toni2009] This dramatically reduces the required compute as multiple model simulations per parameter set are no longer required. The comparison between the observed and simulated data is facilitated using a distance function with parameter sets being accepted if they are within some threshold distance. This threshold can be be tuned to produce a good estimate of the posterior distribution. Developing a distance function for all of the observed data can often be challenging so instead the distance is often calculated using a set of summary features.[@Toni2009] In principle, ABC can be used with a wide variety of algorithms (including a simple rejection approach, MCMC, and SMC) to estimate the posterior distribution but in practice it has been shown that ABC-SMC generally performs better than alternative ABC approaches.[@Toni2009] The two major limitations of ABC compared to the use of a particle filter is that in most cases summary statistics must be used rather than calculating the distance from the observed data and a function for calculating the distance must be chosen.[@Toni2009; @Lintusaari2016] Whilst some techniques exist for evaluating summary statistics the choice is often subjective, relying on domain knowledge.[@Lintusaari2016] Chosen summary statistics rarely capture the information contained in the observed data fully and can inflate posterior distribution and in the worst cases introduce bias to the estimates.[@Busetto2013] The choice of distance function may also influence the estimates of the posterior distribution.[@Lintusaari2016]

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is to complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Alternatively, the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's R interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be frustrating to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards its development. It is possible that the fitting issues outlined here may be due to errors in the code used to implement the model from the previous chapter. This is unlikely as manual prior tuning has shown that a relatively good fit to the observed data can be achieved with the current model but these results could not be reproduced using the model fitting pipeline. However, this does not mean that the modelling code is bug free. Model bugs, if present, would bias results from the fitted model rather than preventing a good fit to the data or would prevent the model from being manually tuned to fit the data.

There are many other tools available within the R ecosystem for fitting infectious disease models and detailing them all is beyond the scope of this thesis. However, `LibBi`, `rbi`, and `rbi.helpers` represent an attempt to provide a complete modelling framework rather than being a simple toolbox for model fitting. The `pomp` package has a similar aim making a comparison worthwhile.[@King2016] `pomp` defines models using a similar structure to that presented here and used within `LiBbi`. However, rather than using its own modelling language it relies on the use of either R or C code. This has several advantages in that `pomp` models can more easily be generalized, can be understood and implemented by users new to the package and can make use of packages from the wider R and C ecosystem. The downside to this approach is that for complex models the use of C is essentially a requirement for efficiency reasons and implementing complex models in C can be an error prone and time consuming process. `pomp` offers support for PMCMC, iterated filtering, and ABC-MCMC but does not support SMC based approaches (such as SMC-SMC and ABC-SMC).[@King2016] In developing the work presented here, and in the previous chapter, a simpler model was developed using `pomp`^[`pomp` modle code: https://gist.github.com/seabbs/f08a8a46b1342b8649df963ac015ea31]. This work was not included in this thesis as it was not sufficiently developed. However, it highlight similar issues with the `pomp` package to those observed when using `LibBi`. Whilst `pomp`'s documentation, stability, and testing were much improved over `LibBi` it had similar limitations when it came to complex models. This was to such an extent that I developed numerous helper functions to deal with both the input and output of `pomp` models^[idmodelr: https://github.com/seabbs/idmodelr]. `pomp`'s documentation also has a heavy focus on iterated filtering over other model fitting techniques.

@@

* Defined the disease transmission model from the previous chapter as a state-space model, outlined the available data to be used for fitting it, and detailed a measurement model to link the observed data with the dynamic TB model.
* Developed a model fitting pipeline based on SMC-SMC to fit the previously defined state-space model to TB notification data. The theoretical background for this approach was outlined as well as the key steps for implementing it in practise. 
* As the quality of model fit achieved was poor ad-hoc calibration approaches, and the results they gave, were discussed.
* The scenarios outlined in the previous chapter were evaluated using DIC and the implications of the findings were discussed. 
* Model forecasts from the best fitting scenario, as established using the DIC, were compared to observed data. Posterior distributions from this model fit were then contrasted with the prior distributions. Finally, parameter sensitivity was estimated using the posterior distributions.
* Discussed the strengths and limitations of the work presented here, as well as outlining potential further work.

>>>>

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding it's structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occurring in England. In order to use this model to understand TB transmission, and the impact of different BCG vaccination policies (Chapter \@ref(model-results)) this uncertainty needs to be reduced and the parameter space tightened to reflect more realistic ranges. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantitative way "better" than other parameter sets. An alternative to model fitting is using a model parameterised with expert knowledge only for inference. This approach is not appropriate here due to the large amount of uncertainty for many of the model parameters. Any inference based on just the parametisation from the previous chapter would have large credible intervals, reflect reality poorly, and likely be biased in multiple areas.

This chapter details an approach to fitting a infectious disease model to data using the state-space model formulation and Bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000. A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre- and post- the introduction of the ETS. However, this may result in over-fitting and so has not been implemented here.

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging from tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates (or interval estimates).[@Murray2015] Unfortunately, many of these methods also require tuning prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to fit the previously detailed state space model.

@@

 * *Re-sampling:* The particle stocke is restorted to equal weights by re-sampling particles, with replacement, with the probability of each sample being drawn being proportional to it's weight.

@@

    *  *Re-sampling:* The parameter particle stock is restored to equal weights by re-sampling particles, with replacement, with the probability of each sample being drawn being proportional to it's weight.

@@

The accuracy of the marginal likelihood estimate returned by the particle filter is dependent on the number of particles used, the number of observed data points, the parameter sample, and the complexity of the model. As the number of particles tends towards infinity the likelihood estimate provided by the particle filter should tend towards the exact solution. This suggests that choosing a very high number of particles may be the most efficient solution in terms of accuracy. Unfortunately, each particle requires a full model simulation, which for complex models can be computationally costly. This means that using very large numbers of particles is not tractable. For this reason it is necessary to determine an optimal number of particles that both provides an adequately accurate estimate of the likelihood whilst being computationally tractable.

@@

The DIC has two limitations. The first of these is that in it's derivation it is assumed that the model that generates future observations encompasses the true model. This assumption may not hold in all circumstances. The second limitations is that the observed data is used to construct both the posterior distribution and to estimate the DIC. This means that the DIC tends to select for over-fitted models.[@Gelman:nll_LBlw]

@@

Where $\operatorname {cov}$  is the covariance, $\sigma _{X}$ is the standard deviation of $X$, and $\sigma_Y$  is the standard deviation of $Y$. Where $X$ is the input and $Y$ is the output.

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC-SMC algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvenation steps in all scenarios evaluated. This resulted in spuriously tight posterior distributions. Ad hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and indicative only.

### Ad hoc calibration

Minor alterations to the model fitting pipeline had little impact on the quality of the fit. To attempt to improve the quality of the model fit I used a combination of ad hoc approaches. As a first step I introduced a calibration model with variation allowed only in the fitted parameters with all other parameters using point estimates. This allowed a reduced number of particles to be used to estimate the marginal likelihood and hence dramatically decreased compute cost and run-times. This reduced model was then used for the following tests:

@@

Murray et al. suggests that increasing the number of particles used in SMC may improve the quality of the model fit.[@Murray2015] Unfortunately as `LibBi` stores SMC particle paths in Random access memory (RAM) the number of particles was restricted. An additional limitation is that the current rejuvenation step also need to be stored in RAM. Attempts to increase the amount of available RAM (64 GB) using a 500GB SWAP (virtual memory) drive increased the upper limit on the number of particles but gains from this were restricted due to thrashing^[Thrashing: https://en.wikipedia.org/wiki/Thrashing_(computer_science)]. Thrashing occurs when to much data is written to SWAP memory in a short period of time and usually results in a system crash.  A major cause of the high RAM requirements of the model was that `LibBi` stores all parameters defined as initial conditions across every time point in the model, regardless of the settings used. Attempting to reduce the RAM footprint by fitting a greater number of parameters allowed for a greater number of particles to be used but also increased the model degrees of freedom and hence upped the required number of particles required by a greater amount than provided by fitting the parameter. Reducing the number of particles used to estimate the marginal likelihood allowed a greater number of particles to be used in the outer SMC step but this resulted in highly inaccurate estimates of the likelihood. These highly inaccurate likelihood estimates resulted in the SMC-SMC algorithm focusing on parameter sets that had a low likelihood estimate yet fit the data poorly - ultimately leading to poorer model fits.

Varying the proposal size, rejuvenation threshold, and number of rejuvenation steps showed some promise at improving the quality of the model fit but ultimately computational constraints limited how much progress could be made using this approach. It is possible that a much longer run time could result in an improved fit to data with no other model changes. Alternatively these results may be driven by the model being too complex for an SMC-SMC approach to be viable.

@@

Whilst none of the scenarios fitted the data well, scenarios that included variable transmission probability in young adults fitted the data much better than those that did not (Table \@ref(tab:09-scenarios-table)). When considered on it's own, allowing non-UK born mixing to vary for young adults had only a small impact on the quality of the model fit in comparison to allowing the transmission probability to vary. However, the scenario that allowed both the transmission probability and non-UK born mixing to vary fit the data much better than any other scenario considered (Table \@ref(tab:09-scenarios-table)). In the following sections only the results from this scenario will be discussed.

@@

```{r 09-table-incidence-preds, message = FALSE, warning = FALSE}
sum_obs_table %>% 
  set_names(., nm = c(names(.)[-ncol(.)], "Predicted Incidence (95% CrI)")) %>% 
   set_names(., nm = names(.) %>% str_replace_all("Incidence", "Cases")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Observed versus predicted overall TB cases for years that the model was fitted to.", 
  caption = "Observed versus predicted overall TB cases for years that the model was fitted to. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  {.}
```

@@

```{r 09-prior-posterior, warning = FALSE, message = FALSE}
sum_prior_posterior %>% 
  set_names(., nm = . %>% 
              str_replace("or", "or (95\\\\%CrI)")) %>%
  mutate(Parameter = latex_param) %>% 
  select(-latex_param) %>% 
kable(
  caption.short = "Prior and posterior medians for fitted model parameters", 
  caption = "Prior and posterior medians for fitted model parameters. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.",
  booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(latex_options = c("HOLD_position")) %>% 
  {.}
```

@@

In this chapter I have formulated the disease transmission model developed in the previous chapter as a state-space model, developed a model fitting pipeline to fit this model to observed data, discussed the approaches taken to try and improve the quality of the model fit, and presented preliminary results from the model fitting pipeline. The model fit the observed data poorly using the approach layed out here. Whilst multiple ad hoc approaches were used to try and improve the quality of the model fit these did little to improve it. The model consistently under-predicted overall TB incidence and also failed to reproduce the observed age distribution. There was little evidence of a good fit to trends over time in the observed data. Model comparison showed that models that included modifying parameters for the transmission probability, and non-UK born mixing in young adults fitted the observed data better than those that did not. The estimated posterior distributions for all parameters were spuriously tight in comparison to the prior distributions. This may be evidence of poor mixing meaning that any conclusions drawn about the parameter sets selected may be incorrect. Parameters that contributed to recent TB transmission dominated those that contributed across the time-span of the model in the parameter sensitivity analysis.

As discussed earlier in this chapter, SMC-SMC was used for all model fitting. This uses an external SMC step to estimate the posterior distribution as well as an internal SMC step to estimate the marginal likelihood. In general, Bayesian model fitting approaches are beneficial as they allow prior information to be fully incorporated into model fitting and they produce posterior distribution estimates rather than point estimates.[@Murray2015; @Toni2009] There are two main families of approaches, SMC and MCMC. Using SMC to estimate the posterior distribution has numerous advantages over MCMC. The first of these is that MCMC approaches are sensitive to their initial conditions. If a model has multiple local best fits MCMC may only converge to a single minima rather than fully exploring the posterior distribution. Multiple MCMC chains may be used to try and account for this but as each chain must be independently run to convergence only a few concurrent chains are likely to be practical. SMC on the other hand is initialised with a large sample from the prior distribution, meaning that local minimas are more likely to be explored. Parameter particle weighting and re-sampling then balances the contribution to the posterior distribution of these local minimas based on their fit to the observed data. Additionally, MCMC approaches are by definition sequential,[@Murray2015] although if they make use of particle filters these can be run in parallel. Increasing the number of particles in a filter may lead to an increase in the chain mixing rate of the MCMC chain but as particles numbers are increased any returns will decrease. To account for this multiple chains are often used, but as outlined above the burn-in required for each chain limits the potential speed-up. In comparison, each SMC parameter particle can have it's marginal likelihood computed separately. Although the re-sampling step remains a bottleneck as it can only be completed once all marginal likelihoods have been computed. On the other hand SMC is less interactive than MCMC meaning that model fitting is harder to inspect when it is in progress. This is because SMC cannot be inspected sequentually, unlike an MCMC run for which each draw can be inspected as it is computed. Similarly, as SMC is not a sequential technique multiple runs cannot be combined. This means that model fitting must be done in a single run using a priori knowledge to judge the number of MCMC rejuvenation steps required, and the expected total run time. SMC will also have a variable run time based on the effective sample size as rejuvenation only happens when parameter particles have been depleted beyond a certain point. An additional benefit of SMC approaches are that they can theoretically be extended to model selection as well as parameter posterior estimation - effectively estimating posteriors for candidate model structures.[@Toni2009] Beyond SMC and MCMC there are multiple other model fitting algorithms that each have their own strengths and benefits - discussion of these is outside the scope of this thesis.

The particle filter has been shown to provide an unbiased estimate of the likelihood for arbitrary state-space models.[@Murray2015] As a full information technique the particle filter provides a more accurate estimate of the likelihood than other approximate techniques with relatively little tuning or user interaction.[@Murray2015] The major downside of the particle filter is the high compute requirements, with each particle requiring a full model simulation. For highly complex models, the particle filter approach may not be tractable or a reduced level of accuracy of the marginal likelihood estimate must be accepted. In addition, the bootstrap particle filter may become depleted (i.e variance between particles is reduce to such an extent that the effective sample size becomes small) when the model being fitted is a poor fit for the observed data.[@Murray2015] Whilst this can be resolved using additional particles, or rejuvenation steps, this may not be computationally tractable. There are several alternative particle filters that seek to mitigate these issues but many of them are highly complex and do not significantly reduce the required compute - see [@Murray2015] for a detailed discussion of some of these alternatives. An alternative approach to the particle filter is to use an approximate technique, such as approximate Bayesian computation (ABC). ABC can be used to avoid having to estimate the likelihood by comparing observed and simulated data.[@Toni2009] This dramatically reduces the required compute as multiple model simulations per parameter set are no longer required. The comparison between the observed and simulated data is facilitated using a distance function with parameter sets being accepted if they are within some threshold distance. This threshold can be be tuned to produce a good estimate of the posterior distribution. Developing a distance function for all of the observed data can often be challenging so instead the distance is often calculated using a set of summary features.[@Toni2009] In principle, ABC can be used with a wide variety of algorithms (including a simple rejection approach, MCMC, and SMC) to estimate the posterior distribution but in practice it has been shown that ABC-SMC generally performs better than alternative ABC approaches.[@Toni2009] The two major limitations of ABC compared to the use of a particle filter is that in most cases summary statistics must be used rather than calculating the distance from the observed data and a function for calculating the distance must be chosen.[@Toni2009; @Lintusaari2016] Whilst some techniques exist for evaluating summary statistics the choice is often subjective, relying on domain knowledge.[@Lintusaari2016] Chosen summary statistics rarely capture the information contained in the observed data fully and can inflate posterior distribution and in the worst cases introduce bias to the estimates.[@Busetto2013] The choice of distance function may also influence the estimates of the posterior distribution.[@Lintusaari2016]

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is too complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Alternatively, the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's R interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be frustrating to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards it's development. It is possible that the fitting issues outlined here may be due to errors in the code used to implement the model from the previous chapter. This is unlikely as manual prior tuning has shown that a relatively good fit to the observed data can be achieved with the current model but these results could not be reproduced using the model fitting pipeline. However, this does not mean that the modelling code is bug free. Model bugs, if present, would bias results from the fitted model rather than preventing a good fit to the data or would prevent the model from being manually tuned to fit the data.

There are many other tools available within the R ecosystem for fitting infectious disease models and detailing them all is beyond the scope of this thesis. However, `LibBi`, `rbi`, and `rbi.helpers` represent an attempt to provide a complete modelling framework rather than being a simple toolbox for model fitting. The `pomp` package has a similar aim making a comparison worthwhile.[@King2016] `pomp` defines models using a similar structure to that presented here and used within `LiBbi`. However, rather than using it's own modelling language it relies on the use of either R or C code. This has several advantages in that `pomp` models can more easily be generalized, can be understood and implemented by users new to the package and can make use of packages from the wider R and C ecosystem. The downside to this approach is that for complex models the use of C is essentially a requirement for efficiency reasons and implementing complex models in C can be an error prone and time consuming process. `pomp` offers support for PMCMC, iterated filtering, and ABC-MCMC but does not support SMC based approaches (such as SMC-SMC and ABC-SMC).[@King2016] In developing the work presented here, and in the previous chapter, a simpler model was developed using `pomp`^[`pomp` modle code: https://gist.github.com/seabbs/f08a8a46b1342b8649df963ac015ea31]. This work was not included in this thesis as it was not sufficiently developed. However, it highlight similar issues with the `pomp` package to those observed when using `LibBi`. Whilst `pomp`'s documentation, stability, and testing were much improved over `LibBi` it had similar limitations when it came too complex models. This was to such an extent that I developed numerous helper functions to deal with both the input and output of `pomp` models^[idmodelr: https://github.com/seabbs/idmodelr]. `pomp`'s documentation also has a heavy focus on iterated filtering over other model fitting techniques.

@@

* Defined the disease transmission model from the previous chapter as a state-space model, outlined the available data to be used for fitting it, and detailed a measurement model to link the observed data with the dynamic TB model.
* Developed a model fitting pipeline based on SMC-SMC to fit the previously defined state-space model to TB notification data. The theoretical background for this approach was outlined as well as the key steps for implementing it in practise. 
* As the quality of model fit achieved was poor ad hoc calibration approaches, and the results they gave, were discussed.
* The scenarios outlined in the previous chapter were evaluated using DIC and the implications of the findings were discussed. 
* Model forecasts from the best fitting scenario, as established using the DIC, were compared to observed data. Posterior distributions from this model fit were then contrasted with the prior distributions. Finally, parameter sensitivity was estimated using the posterior distributions.
* Discussed the strengths and limitations of the work presented here, as well as outlining potential further work.
